Haskell in Computational Biology

### 2013-05-27 hov4.tex
build instructions

vee' Ins (NC 0) i = intoInsZero i
vee' Mat (NC 1) i = intoMatOne i
vee' Del (NC 1) i = intoDelOne i
vee' stateHat j i =
  internal [ edge score state (vee'' state pj pi)
           | state <- preceders stateHat
           , let pi = case state of { Del -> i ; _ -> pred i }
           , pj >= 0, pi >= 0
           , let score = transition (node pj) state stateHat
                         + emission (node pj) state (residue pi)
           ]
  where pj = case stateHat of { Ins -> j ; _ -> pred j }

>>>>

vee' :: StateLabel -> NodeIndex -> ResidueCount -> a
vee' Ins (NI 0) i = intoInsZero i
vee' Mat (NI 1) i = intoMatOne i
vee' Del (NI 1) i = intoDelOne i
vee' stateHat j i =
  internal [ edge score state (vee'' state pj pi)
           | state <- preceders stateHat
           , let pi = case state of { Del -> i ; _ -> pred i }
           , pj >= 0, pi >= 0
           , let score = transition (node pj) state stateHat
                         + emission (node pj) state (residue pi)
           ]
  where pj = case stateHat of { Ins -> j ; _ -> pred j }

### 2013-05-27 jfp.tex
build instructions

Our first step in simplification was to start thinking like Haskell
programmers and use types:
\begin{itemize}
\item
We~refactored the single, flat \texttt{Node} type from our ICFP paper
into the three state types you see on page~\pageref{code:model3-node}.
\item 
We introduced a \texttt{Model} type that consists of a special Begin
node containing a begin and an insertion state, an~unbounded sequence
of ordinary \texttt{Nodes}, and an end state.
\item
We~observed that the main algorithm where \mrfy\ spends all its time
uses only match, insertion, and deletion states, so we removed the
value constructors corresponding to begin and end states.
Those constructors were used only to prettyprint models, so~we made
them part of a larger type.
\end{itemize}
As a result of these changes, in the revised code,
\begin{itemize}
\item
Only array index~0 is used to indicate a special
condition: the transition from the leftmost middle node to the unique
begin node.
\item
All well-typed combinations of array indices and states make sense;
senseless combinations are ruled out by the static type system.
\end{itemize}

@@

Our next step was to use more abstraction.
In~Viterbi's equations,
$V'$ is indexed by a residue index~$i$,\footnote
{In~biology, when an amino acid enters a protein, it often loses a few
atoms, and so within the protein, the amino acid is often called
a \emph{residue}.}
 a~node index~$j$,
and
a state $M$, $I$, or~$D$.
Each state uses a slightly different equation, but it is possible to
combine them.
\figref{viterbi-abstracted} shows a single equation which, when
state $\hat s$ is specialized to state $M$, $I$, or~$D$, recovers the
original three equations from \figref{viterbi-transformed}.
The figure sums the cost of emitting an amino acid at state~$\hat s$
lowest of the costs of the paths entering~$\hat s$ from predecessor
state~$s$.
Index~$\p j$ is the index of the node containing the predecessor
state, and $\p i; s$ is the index of the amino acid, if~any, to be
emitted at state~$s$.

NEED TO EXPLAIN PRECEDERS

\figref{viterbi-abstracted} doesn't deal with the begin state.
In~our legacy file format, and also in our code, the begin state is
represented as a match state inside an ordinary node.
In~a perfect world, then, we would add one case to say that the match
state with $i=j=0$ is reachable at zero cost, and we'd be finished.
But the world is imperfect, because although the begin state is
represented as a match state, unlike a match state, the begin state
does not emit.
We~therefore need a special case for each of the three states that can
follow the begin state.
Here is the code:
\smallverbatiminput{hov4}
NEED TO EXPLAIN INTERNAL, EDGE, AND LEAF.
NEED TO EXPLAIN [[transition]] and [[emission]]

>>>>

Our first step in simplification was to start thinking like Haskell
programmers and use types:
\begin{itemize}
\item
We~refactored the single, flat \texttt{Node} type from our ICFP paper
into the three state types you see on page~\pageref{code:model3-node}.
\item 
We introduced a \texttt{Model} type that consists of a special Begin
node containing a begin and an insertion state, an~unbounded sequence
of ordinary \texttt{Nodes}, and an end state.
And we represented the sequence as a list, not an array.
\item
We~observed that the main algorithm where \mrfy\ spends all its time
uses only match, insertion, and deletion states, so we removed the
\texttt{StateLabel} value constructors corresponding to begin and end
states, leaving only the three constructors shown on
page~\ref{code:state-label}. 
(We~moved the begin and end labels to another type, which is used only
to prettyprint models.) 
\end{itemize}
The immediate effect of these changes was to reduce case analysis on
two dimensions:
instead of five state labels, we got three;
and instead of three special array indices, we got just one special
list (the empty list).

\begin{figure}
\noindent\hspace*{1.5cm}\begin{minipage}{9cm}
\smallverbatiminput{list-viterbi}
\smallskip
\end{minipage}

\caption{Sketch of Viterbi's algorithm, using lists}
\figlabel{list-viterbi}
\end{figure}

Our~next step was to use more abstraction---in~particular, to abstract
over state labels.
However, after abstracting over state labels, we~abandoned lists and
went back to arrays.
An~excerpt from our list-based code appears in \figref{list-viterbi}.
The~problem is defining \verb+vee''+ as a memoized version
of~\verb+vee'+.
Although we experimented with specialized memo combinators that work
with list prefixes, 
we~were never confident that we could easily, efficiently, and
correctly memoize~\verb+vee'+.
Instead, we~wrote a new version, still using integer arithmetic, but
guides by the idea $i$~is a \emph{count} of the number
of amino acids in list~\texttt{aas}, but $j$~is the \emph{index}
of~\texttt{node}.
(We~also got
an
insight we got from
the \texttt{next} function: Viterbi's algorithm is a form of minimum edit
distance.
As~we were trying to improve the code, this insight helped keep us on track.)
Armed with this new, better way of thinking about $i$~and~$j$,
we~turned back to arrays.

@@

Our final code, like the code in \figref{list-viterbi}, is abstracted
over state labels.
We~now turn your attention to that abstraction.
In~Viterbi's equations,
$V'$ is indexed by a residue count~$i$,\footnote
{In~biology, when an amino acid enters a protein, it often loses a few
atoms, and so within the protein, the amino acid is often called
a \emph{residue}.}
a~node index~$j$,
and
a state $M$, $I$, or~$D$.
Each state uses a slightly different equation, but it is possible to
combine them.
\figref{viterbi-abstracted} shows a single equation which, when
state $\hat s$ is specialized to state $M$, $I$, or~$D$, recovers the
original three equations from \figref{viterbi-transformed}.
The figure sums the cost of emitting an amino acid at state~$\hat s$
lowest of the costs of the paths entering~$\hat s$ from predecessor
state~$s$.
Index~$\p j$ is the index of the node containing the predecessor
state, and $\p i; s$ is the index of the amino acid, if~any, to be
emitted at state~$s$.

To~implement \figref{viterbi-abstracted}, we~define several
auxiliary functions.
The~set of states that can precede~$\hat s$
is \mono{preceders~$\hat s$}; $\p i$~and~$\p j$ are computed using
function \texttt{predUnless}:
\smallverbatiminput{v4aux}
We~also use \texttt{newtype} to define 
a value constructor~\texttt{RC} for a residue count~$i$
and
a value constructor~\texttt{NI} for a node index~$j$.
We~can then deliver the more abstract version of Viterbi's algorithm
shown in \figref{viterbi-hov4}.

\begin{figure}
\noindent\hspace*{1.0cm}\begin{minipage}{11cm}
\smallverbatiminput{hov4}
\smallskip
\end{minipage}

\caption{Viterbi's algorithm abstracted over~$\hat s$}
\figlabel{viterbi-hov4}
\end{figure}

The version of \verb+vee'+ shown in  \figref{viterbi-hov4} is abstract
in another way: the type of the result is the unknown type~\texttt{a}.
Type~\texttt{a} represents

\figref{viterbi-abstracted} doesn't deal with the begin state.
In~our legacy file format, and also in our code, the begin state is
represented as a match state inside an ordinary node.
In~a perfect world, then, we would add one case to say that the match
state with $i=j=0$ is reachable at zero cost, and we'd be finished.
But the world is imperfect, because although the begin state is
represented as a match state, unlike a match state, the begin state
does not emit.
We~therefore need a special case for each of the three states that can
follow the begin state.

Here is the code:
NEED TO EXPLAIN INTERNAL, EDGE, AND LEAF.
NEED TO EXPLAIN [[transition]] and [[emission]]

As a result of these changes, in the revised code,
\begin{itemize}
\item
Only array index~0 is used to indicate a special
condition: the transition from the leftmost middle node to the unique
begin node.
\item
\end{itemize}

\begin{verbatim}
### 2013-05-27 01:23   Running med(vpasses=1): icfp-branch mrfy nr-jfp
icfp-branch faab8dc       54.07s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
mrfy d5604bc*             44.90s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
nr-jfp 00ef54e*           47.59s    i686  GHC 7.4.1  Core i5-3570 @ 3.40GHz
\end{verbatim}
